A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-09-27 paper.tex
minor changes

% random stuff
\newcommand{\infr}[2]{\inference{#1}{#2}}
\newcommand{\highlight}[1]{{\color{red} #1}}
\newcommand{\sidecond}[1]{\text{if}\;#1}

@@

Gradual typing \citep{siek2006gradual,siek2015refined} includes both static typing 
and dynamic typing in the same language, enables smooth transition between them, while
enforcing type soundness.
To allow the smooth transition, gradually typed languages are featuerd by a special type
typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|, which enables 
imprecise type annotations. A gradual type checker allows programmers to cast 
implicitly between types that have no conflicting type information.
Researchers usually define the semantics of gradually typed languages by
translation to an intermediate language with explicit casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi \citep{siek2009exploring}.
At runtime, casts store or check type information to enforce type soundness.
% To allow the smooth transition, every gradually typed language
% has a special type typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|,
% which enables imprecise type annotations.
% Programmers can use an expression of type \texttt{T} in any context that \texttt{T} 
% is consistent to.
% Two types are consistent if they have no conflicting type information.
% For instance, \lstinline|(->  Bool Int)| is consistent with \lstinline|(->  $\star$ Int)| but
% not consistent with \lstinline|(->  Int Int)|.
% When the type of an expression is consistent but not equal to the type of its 
% context, there would be a runtime computation (i.e. a cast) that store or check 
% type information to enforce type soundness.

@@

Of course, an alternative cast representation must be proved
correct. That is, a language implementation should give the same output no 
matter whether it is using a cast representation to reduce space 
consumption. Our generic CEK machine comes with generic theorems (Section~\ref{sec:framework}),
with which researchers can prove a cast representation is correct by showing its corresponding
implementation of Cast ADT satisfies a bunch of properties. We believe this framework can reduce the burden
 of proving this kind of theorems and help people focusing on the cast representations.
For illustration, we apply the framework to Lazy UD coercions in normal form
and Lazy UD threesomes. 
We believe the framework is also applicable to other cast representations.
\todo[inline]{KC: (about the p above) I want to put the illustration part on the same github repo, adding description as
Agda comments. }

@@

Next, we define the CEK Machine for \LUD. The only differences with
respect to \LD\ are in the injectable types and in the definition of the 
${\mathit{applyCast}(v,\ccast{S}{l}{T})}$
function, in which a cast whose source or target is the dynamic type
\Tdyn\ is always split into two casts that go through a ground type, that is, a 
type in which all sub-components are the dynamic type, such as 
\Pfunc{\Tdyn}{\Tdyn}.
\begin{definition}[\LUD\ CEK Machine]
	The \LUD\ CEK machine, written \LUDMachine\, is the CEK machine of
	\figref{fig:R-machine-dynamics} using the \BLUD\ blame strategy defined in 
	\figref{fig:blame-strategy-LazyUD}.  We write the transition relations of
	this machine as \reducetoUD{s}{s} and \transitivelyreducetoUD{s}{s}
	and the evaluator as \evaltoUD{e}{o}.
\end{definition}

@@

\figref{fig:semachine-dynamics} defines the transition relation 
$\reducetoS{C}{s}{s}$. They rely on functions provided 
by $C$ to work with casts.
%
To construct a function value, the machine fills the casts with identities.
%
When evaluating a cast expression, the machine extends the continuation with 
\translate{\ccast{S}{l}{T}}.
%
To return a value to a continuation, the machine first applies the top cast to 
the value. If the cast application fails, the machine halts with the blame 
label from the failure. Otherwise, the machine handles the pre-continuation 
with the function $\applyCont{v}{k}$.
%
To apply a function, the machine applies the domain cast $c$ to the 
argument $v$ and evaluates the function body. The codomain cast $d$ is merged 
into the continuation with $\mathrm{extK}$.
%
When the machine constructs a pair, it fills the casts with identities, just 
like how it does for function values. 
%
To take out the first part of a pair, the machine applies the first pending 
cast on the first value and returns the result to the continuation.
%
Taking out the second part of a pair is similar.
\todo[inline]{KC: I want to give a brief discussion on inert v.s. active pairs.}

@@

\todo[inline]{KC: In the latex file, below are two versions of a figure.
One is the same as in our WGT paper. The other uses Z as the meta variable of sequences of 
source-code casts, which is similar to the blame-and-coercion paper.}

@@

\fbox{$\bisimv{v}{v}{T}$}
\begin{gather*}
\infr{
	\expressiontyping{\Gamma,x : S_1}{e}{T_1}
	\qquad
	\bisimE{\Gamma}{E}{E'}
	\\
	\bisimc{
		(\ccast{S_n}{l_{n-1}}{S_{n-1}} \ldots
		\ccast{S_2}{l_1}{S_1})
	}{c}{S_n}{S_1}
	\\
	\bisimc{
		(		\ccast{T_1}{l_{1}}{T_2} \ldots
		\ccast{T_{n-1}}{l_{n-1}}{T_n}
		)	}{d}{T_1}{T_n}
	\\
	v = \vfuncc{
		\vfuncc{
			\vfunc{x}{S_1}{e}{E}
		}{S_1}{T_1}{l_1}{S_2}{T_2}
		\ldots
	}{S_{n-1}}{T_{n-1}}{l_{n-1}}{S_n}{T_n}
}{
	\bisimv{v}{
		(\ecast{\vfunc{x}{S_1}{e}{E'}}{\Pfunc{c}{d}})
	}{
		\Pfunc{S_n}{T_n}}
}
\\[1ex]
\infr{
	\bisimv{v_1}{v_1'}{S_1} \qquad
	\bisimv{v_2}{v_2'}{T_1} \\
	\bisimc{
		(\ccast{S_1}{l_{1}}{S_2} \ldots
		\ccast{S_{n-1}}{l_{n-1}}{S_n})
	}{c}{S_1}{S_n} \\
	\bisimc{
		(\ccast{T_1}{l_{1}}{T_2} \ldots
		\ccast{T_{n-1}}{l_{n-1}}{T_n})
	}{d}{T_1}{T_n}
	\\ v = \vconsc{
		\vconsc{
			\vcons{v_1}{u_1}
		}{S_1}{T_1}{l_1}{S_2}{T_2}
		\ldots
	}{S_{n-1}}{T_{n-1}}{l_{n-1}}{S_n}{T_n}
}{
	\bisimv{
		v
	}{
		(\ecast{\vcons{v_2}{u_2}}{\Pprod{c}{d}})
	}{
		\Pprod{S_n}{T_n}}
}
\end{gather*}

\fbox{\bisimc{Z}{c}{S}{T}}
\begin{gather*}
\infr{}{\bisimc{[]}{id(T)}{T}{T}} \quad
\infr{}{\bisimc{[\ccast{S}{l}{T}]}{\translate{\ccast{S}{l}{T}}}{S}{T}}
\\[1ex]
\infr{
	\bisimc{Z_1}{c_1}{T_1}{T_2}\\
	\bisimc{Z_2}{c_2}{T_2}{T_3}
}{
	\bisimc{(Z_1 \mathrm{++} Z_2)}{(\compose{c_1}{c_2})}{T_1}{T_3}
}
\end{gather*}

\fbox{$\bisimK{K}{K}{S}{T}$}
\begin{gather*}
\infr{
	\bisimK{K}{k'}{S_n}{T} \qquad
	\bisimc{Z}{c}{S_1}{S_n}
}{
	\bisimK{
		\Kcons{\Fcast{Z}}{K}
	}{
		\Kcons{\Fcast{c'}}{k'}
	}{S_1}{T}
}
\end{gather*}

@@

@@

\fbox{$\composegap{m}{g}{m} = m$}
\[ 
\begin{array}{rclr}
\figequalto{\composegap{\hcmfail{l}}{g}{m_2}}{
	\hcmfail{l}
}{}
\figequalto{\composegap{m_1}{\gsome{P}{l}{Q}}{m_2}}{
	\hcmfail{l}
}{\sidecond{
		\notshallowlyconsistent{P}{Q}
		\text{ and } \forall l. m_1 \neq \hcmfail{l}
}}
\figequalto{\composegap{m_1}{g}{\hcmfail{l}}}{
	\hcmfail{l}
}{\sidecond{
		\goodgap{g}
		\text{ and } \forall l. m_1 \neq \hcmfail{l}}}
\figequalto{\composegap{\hcmbool}{g}{\hcmbool}}{
	\hcmbool
}{}
\figequalto{
	\composegap{
		(\hcmfunc{c_1}{d_1})}{
		g}{
		(\hcmfunc{c_2}{d_2})}
}{
	\hcmfunc{
		(\composegap{c_2}{\dom{g}}{c_1})}{
		(\composegap{d_1}{\cod{g}}{d_2})}
}{}
\figequalto{
	\composegap{
		(\hcmprod{c_1}{d_1})}{
		g}{
		(\hcmprod{c_2}{d_2})}
}{
	\hcmprod{
		(\composegap{c_1}{\fst{g}}{c_2})}{
		(\composegap{d_1}{\snd{g}}{d_2})}
}{}
\end{array}
\]

>>>>

% random stuff
\newcommand{\infr}[2]{\inference{#1}{#2}}
\newcommand{\highlight}[1]{{\color{red} #1}}
\newcommand{\sidecond}[1]{\text{if}\;#1}
\newcommand{\urlFrameworkAndIllustration}{
	\url{https://github.com/LuKC1024/esop-2020}
}

@@

Gradual typing \citep{siek2006gradual,siek2015refined} includes both static typing 
and dynamic typing in the same language, enables smooth transition between them, while
enforcing type soundness.
To allow the smooth transition, gradually typed languages are featuerd by a special type
typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|, which enables 
imprecise type annotations. A gradual type checker allows programmers to cast 
implicitly between types that have no conflicting type information.
At runtime, casts store or check type information to enforce type soundness.
Researchers usually define the semantics of gradually typed languages by
translation to an intermediate language with explicit casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi \citep{siek2009exploring}.
% To allow the smooth transition, every gradually typed language
% has a special type typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|,
% which enables imprecise type annotations.
% Programmers can use an expression of type \texttt{T} in any context that \texttt{T} 
% is consistent to.
% Two types are consistent if they have no conflicting type information.
% For instance, \lstinline|(->  Bool Int)| is consistent with \lstinline|(->  $\star$ Int)| but
% not consistent with \lstinline|(->  Int Int)|.
% When the type of an expression is consistent but not equal to the type of its 
% context, there would be a runtime computation (i.e. a cast) that store or check 
% type information to enforce type soundness.

@@

Of course, an alternative cast representation must be proved
correct. That is, a language implementation should give the same output no 
matter whether it is using a cast representation to reduce space 
consumption. Our generic CEK machine comes with generic theorems (Section~\ref{sec:framework}),
with which researchers can prove a cast representation is correct by showing its corresponding
implementation of Cast ADT satisfies a bunch of properties. We believe this framework can reduce the burden
 of proving this kind of theorems and help people focusing on the cast representations.
For illustration, we apply the framework to Lazy UD coercions in normal form
and Lazy UD threesomes. 
We believe the framework is also applicable to other cast representations.
The framework and the illustration are available at the following URL:
\[\text{\urlFrameworkAndIllustration}\]

@@

Next, we define the CEK Machine for \LUD. The only differences with
respect to \LD\ are in the injectable types and in 
${\mathit{applyCast}(v,\ccast{S}{l}{T})}$, in which a cast whose source or
target is the dynamic type \Tdyn\ is always split into two casts that
go through a ground type, that is, a 
type in which all sub-components are the dynamic type, such as 
\Pfunc{\Tdyn}{\Tdyn}.
\begin{definition}[\LUD\ CEK Machine]
	The \LUD\ CEK machine, written \LUDMachine\, is the CEK machine of
	\figref{fig:R-machine-dynamics} using the \BLUD\ blame strategy defined in 
	\figref{fig:blame-strategy-LazyUD}.  We write the transition relations of
	this machine as \reducetoUD{s}{s} and \transitivelyreducetoUD{s}{s}
	and the evaluator as \evaltoUD{e}{o}.
\end{definition}

@@

\figref{fig:semachine-dynamics} defines the transition relation 
$\reducetoS{C}{s}{s}$. They rely on functions provided 
by $C$ to work with casts.
%
To construct a function value, the machine fills the casts with identities.
%
When evaluating a cast expression, the machine extends the continuation with 
\translate{\ccast{S}{l}{T}}.
%
To return a value to a continuation, the machine first applies the top cast to 
the value. If the cast application fails, the machine halts with the blame 
label from the failure. Otherwise, the machine handles the pre-continuation 
with the function $\applyCont{v}{k}$.
%
To apply a function, the machine applies the domain cast $c$ to the 
argument $v$ and evaluates the function body. The codomain cast $d$ is merged 
into the continuation with $\mathrm{extK}$.
%
When the machine constructs a pair, it fills the casts with identities, just 
like how it does for function values. 
%
To take out the first part of a pair, the machine applies the first pending 
cast on the first value and returns the result to the continuation.
%
Taking out the second part of a pair is similar.

@@

@@

@@

% \begin{figure}

% \fbox{\bisimc{Z}{c}{S}{T}}
% \begin{gather*}
% \infr{}{\bisimc{[]}{id(T)}{T}{T}} \quad
% \infr{}{\bisimc{[\ccast{S}{l}{T}]}{\translate{\ccast{S}{l}{T}}}{S}{T}}
% \\[1ex]
% \infr{
% 	\bisimc{Z_1}{c_1}{T_1}{T_2}\\
% 	\bisimc{Z_2}{c_2}{T_2}{T_3}
% }{
% 	\bisimc{(Z_1 \mathrm{++} Z_2)}{(\compose{c_1}{c_2})}{T_1}{T_3}
% }
% \end{gather*}

% \fbox{$\bisimK{K}{K}{S}{T}$}
% \begin{gather*}
% \infr{
% 	\bisimK{K}{k'}{S_n}{T} \qquad
% 	\bisimc{Z}{c}{S_1}{S_n}
% }{
% 	\bisimK{
% 		\Kcons{\Fcast{Z}}{K}
% 	}{
% 		\Kcons{\Fcast{c'}}{k'}
% 	}{S_1}{T}
% }
% \end{gather*}

% \fbox{$\bisimv{v}{v}{T}$}
% \begin{gather*}
% \infr{
% 	\expressiontyping{\Gamma,x : S_1}{e}{T_1}
% 	\qquad
% 	\bisimE{\Gamma}{E}{E'}
% 	\\
% 	\bisimc{
% 		(\ccast{S_n}{\bar{l}_{n-1}}{S_{n-1}} \ldots
% 		\ccast{S_2}{\bar{l}_1}{S_1})
% 	}{c}{S_n}{S_1}
% 	\\
% 	\bisimc{
% 		(		\ccast{T_1}{l_{1}}{T_2} \ldots
% 		\ccast{T_{n-1}}{l_{n-1}}{T_n}
% 		)	}{d}{T_1}{T_n}
% 	\\
% 	v = \vfuncc{
% 		\vfuncc{
% 			\vfunc{x}{S_1}{e}{E}
% 		}{S_1}{T_1}{l_1}{S_2}{T_2}
% 		\ldots
% 	}{S_{n-1}}{T_{n-1}}{l_{n-1}}{S_n}{T_n}
% }{
% 	\bisimv{v}{
% 		(\ecast{\vfunc{x}{S_1}{e}{E'}}{\Pfunc{c}{d}})
% 	}{
% 		\Pfunc{S_n}{T_n}}
% }
% \\[1ex]
% \infr{
% 	\bisimv{v_1}{v_1'}{S_1} \qquad
% 	\bisimv{v_2}{v_2'}{T_1} \\
% 	\bisimc{
% 		(\ccast{S_1}{l_{1}}{S_2} \ldots
% 		\ccast{S_{n-1}}{l_{n-1}}{S_n})
% 	}{c}{S_1}{S_n} \\
% 	\bisimc{
% 		(\ccast{T_1}{l_{1}}{T_2} \ldots
% 		\ccast{T_{n-1}}{l_{n-1}}{T_n})
% 	}{d}{T_1}{T_n}
% 	\\ v = \vconsc{
% 		\vconsc{
% 			\vcons{v_1}{u_1}
% 		}{S_1}{T_1}{l_1}{S_2}{T_2}
% 		\ldots
% 	}{S_{n-1}}{T_{n-1}}{l_{n-1}}{S_n}{T_n}
% }{
% 	\bisimv{
% 		v
% 	}{
% 		(\ecast{\vcons{v_2}{u_2}}{\Pprod{c}{d}})
% 	}{
% 		\Pprod{S_n}{T_n}}
% }
% \end{gather*}

% \fbox{\bisimE{\Gamma}{E}{E}}
% \begin{gather*}
% \inference{
% }{
% 	\bisimE{\emptyset}{\emptyset}{\emptyset}
% }
% \quad
% \inference{
% 	\Gamma \vdash E \approx E' \qquad
% 	v \approx v' : T
% }{
% 	\Gamma , x : T \vdash 
% 	E [ x := v' ] \approx E' [ x := v' ]
% }
% \end{gather*}

% \fbox{\bisims{s}{s}{T}}
% \begin{gather*}
% \infr{}{\bisims{\error{l}}{\error{l}}{T}} \quad
% \infr{
% 	\bisimv{v}{v'}{T}
% }{
% 	\bisims{\shalt{v}}{\shalt{v'}}{T}
% }
% \\
% \infr{
% 	\bisimv{v}{v'}{S} \\
% 	\bisimK{K}{K'}{S}{T}
% }{
% 	\bisims{\scont{v}{K}}{\scont{v}{K'}}{T}
% } \quad
% \infr{
% 	\expressiontyping{\Gamma}{e}{S} \\
% 	\bisimE{\Gamma}{E}{E'}          \\
% 	\bisimK{K}{K'}{S}{T}
% }{
% 	\bisims{\sexpr{e}{E}{K}}{\sexpr{e}{E'}{K'}}{T}
% }
% \end{gather*}

% \caption{Bisimulation between \RBMachine\ and \SMachine{C}. It is the smallest 
% congruence relation containing the shown rules. Everything to the left (resp. 
% right) of $\approx$ is in \RBMachine\ (resp. in \SMachine{C}).}
% \label{fig:bisim-SC-D}
% \end{figure}

@@

\fbox{$\composegap{m}{g}{m} = m$}
\[ 
\begin{array}{rclr}
\figequalto{\composegap{\hcmfail{l}}{g}{m_2}}{
	\hcmfail{l}
}{}
\figequalto{\composegap{m_1}{\gsome{P}{l}{Q}}{m_2}}{
	\hcmfail{l}
}{\\&&\sidecond{
		\notshallowlyconsistent{P}{Q}
		\text{ and } \forall l. m_1 \neq \hcmfail{l}
}}
\figequalto{\composegap{m_1}{g}{\hcmfail{l}}}{
	\hcmfail{l}
}{\\&&\sidecond{
		\goodgap{g}
		\text{ and } \forall l. m_1 \neq \hcmfail{l}}}
\figequalto{\composegap{\hcmbool}{g}{\hcmbool}}{
	\hcmbool
}{}
\figequalto{
	\composegap{
		(\hcmfunc{c_1}{d_1})}{
		g}{
		(\hcmfunc{c_2}{d_2})}
}{
	\hcmfunc{
		(\composegap{c_2}{\dom{g}}{c_1})}{
		(\composegap{d_1}{\cod{g}}{d_2})}
}{}
\figequalto{
	\composegap{
		(\hcmprod{c_1}{d_1})}{
		g}{
		(\hcmprod{c_2}{d_2})}
}{
	\hcmprod{
		(\composegap{c_1}{\fst{g}}{c_2})}{
		(\composegap{d_1}{\snd{g}}{d_2})}
}{}
\end{array}
\]
